---
title: "community_analysis_8.20"
author: "Scott Klasek"
date: "8/26/2020"
output: github_document
---

## load necessary libraries and data
```{r}
library(phyloseq)
library(tidyverse)
library(egg)
library(lemon)
library(gridExtra)
library(grid)
library(patchwork)
library(vegan)
library(DESeq2)
library(here)
sessioninfo <- sessionInfo()
ps.frdp <- readRDS(file="ps.frdp") # imports the final phyloseq object
```

## alpha diversity
```{r}
plotrichness.oc <- plot_richness(ps.frdp, x="dist_above_peakaom", measures=c("Shannon"), color="core")
plotrichness.oc+
  facet_grid(core_flowtype~.)+
  ylab("Shannon diversity index")+
  xlab("Distance above present-day peak AOM rate (cm)")+
  scale_color_discrete("Core")+
  theme_bw()

ps.frdp.ss <- subset_samples(ps.frdp, sample_data(ps.frdp)$stage=="steadystate") # subsetsall steady-state samples
ps.frdp.nss.seep <- subset_samples(ps.frdp, sample_data(ps.frdp)$stage!="steadystate") # subsets the non-steady-state samples and the seep

alphadiv.ss <- estimate_richness(ps.frdp.ss, split = TRUE, measures=c("Shannon","InvSimpson","Observed")) # calculate tables of alpha diversity
alphadiv.nss <- estimate_richness(ps.frdp.nss.seep, split = TRUE, measures=c("Shannon","InvSimpson","Observed"))

alphadiv.ss$depth <- substring(rownames(alphadiv.ss), 8,12) # pull out depth values from the sample names
alphadiv.ss$depth <- as.numeric(alphadiv.ss$depth) # convert them to numeric
alphadiv.nss$depth <- substring(rownames(alphadiv.nss), 8,12)
alphadiv.nss$depth <- as.numeric(alphadiv.nss$depth)

alphadiv.ss$core <- substring(rownames(alphadiv.ss), 1,6) # pull out core names from sample names
alphadiv.nss$core <- substring(rownames(alphadiv.nss), 1,6)

alphadiv.nss$peakaom <- NA # creates a NA value for peak AOM depth
alphadiv.ss$peakaom <- NA
alphadiv.nss[which(alphadiv.nss$core=="GC1081"),6] <- 56 # add in SMT depths for all cores
alphadiv.nss[which(alphadiv.nss$core=="GC1045"),6] <- 65
alphadiv.nss[which(alphadiv.nss$core=="PC1029"),6] <- 45.3
alphadiv.ss[which(alphadiv.ss$core=="GC1068"),6] <- 108
alphadiv.ss[which(alphadiv.ss$core=="GC1069"),6] <- 137
alphadiv.ss[which(alphadiv.ss$core=="GC1070"),6] <- 67
alphadiv.ss[which(alphadiv.ss$core=="GC1048"),6] <- 313.5
alphadiv.ss$dist_above_peakaom <- alphadiv.ss$peakaom-alphadiv.ss$depth # calculate distance above/below SMT
alphadiv.nss$dist_above_peakaom <- alphadiv.nss$peakaom-alphadiv.nss$depth 
alphadiv.ss.range <- subset(alphadiv.ss, dist_above_peakaom < 62 & dist_above_peakaom > -46) # subset the dataframe to include steady-state samples from depths across the SMT that correspond to depths of non-steady-state samples

nss.lm <- lm(alphadiv.nss$Shannon~alphadiv.nss$dist_above_peakaom) # linear model of shannon alpha diversity over distance to SMT for non-steady-state cores
ss.lm <- lm(alphadiv.ss$Shannon~alphadiv.ss$dist_above_peakaom) # linear model of shannon alpha diversity over distance to SMT for non-steady-state cores
ss.lm.range <- lm(alphadiv.ss.range$Shannon~alphadiv.ss.range$dist_above_peakaom)
summary(nss.lm) # summarizes the linear regression: slope is 0.023203, intercept is 3.777779, both are significant. Multiple R^2 is 0.4874, p value of slope is 5.14e-05
summary(ss.lm) # slope is significant considering all the points across depth 
summary(ss.lm.range) # slope is 0.006184, intercept is 3.742064, slope not significant across the SMT range. Multiple R^2 is 0.03277, p value of slope is 0.347

# assign facet grid labels and methane flux stages
alphadiv.nss$core_flowtype_label <- "High Methane Flux"
alphadiv.ss$core_flowtype_label <- "Low Methane Flux"
alphadiv.nss$stage <- "increasing methane flux"
alphadiv.nss[which(alphadiv.nss$core=="PC1029"),9] <- "active methane seepage"
alphadiv.ss$stage <- "steady-state"

plotrichness.ss <- ggplot(alphadiv.ss, aes(dist_above_peakaom, Shannon, color=stage, shape=core)) 
ssplot <- plotrichness.ss+
  geom_point(size=2)+
  facet_grid(~core_flowtype_label)+
  scale_y_continuous("Shannon diversity index", breaks=c(1,2,3,4,5,6), limits = c(0.8,6.5))+
  scale_x_continuous("Distance above present-day peak AOM rate (cm)", limits = c(-250,306))+
  scale_color_manual("Methane stage", values=c("#8da0cb"))+
  scale_shape_manual("Core",values = c(4:7))+
  geom_vline(aes(xintercept=-45), linetype="dashed", size=0.8)+
  geom_vline(aes(xintercept=61), linetype="dashed", size=0.8)+
  geom_segment(x=-45,y=3.482369,xend=61,yend=4.116885, color="black", size=1)+
  theme_bw()+ # now for some reason you need to put the theme_bw() ahead of the theme() or nothing in theme() will work
  theme(strip.text = element_text(size = 11))+
  annotate("text", x = -150, y = 4.9, label = "paste(italic(R) ^ 2, \" = 0.0328\")", parse = TRUE)+
  annotate("text", x = -150, y = 5.6, label = "paste(italic(p), \" = 0.347\")", parse = TRUE)+
  annotate("rect", xmin = -225, xmax = -75, ymin = 4.5, ymax = 6, alpha = .2)

plotrichness.nss <- ggplot(alphadiv.nss, aes(dist_above_peakaom, Shannon, color=stage, shape=core))
nssplot <- plotrichness.nss+
  geom_point(size=2)+
  facet_grid(~core_flowtype_label)+
  scale_y_continuous("Shannon diversity index", breaks=c(1,2,3,4,5,6), limits = c(0.8,6.5))+
  scale_x_continuous("Distance above present-day peak AOM rate (cm)", limits = c(-250,306))+
  scale_color_manual("Methane stage", values=c("#fc8d62", "#66c2a5"))+
  scale_shape_manual("Core", values = c(1:3))+
  geom_vline(aes(xintercept=-45), linetype="dashed", size=0.8)+
  geom_vline(aes(xintercept=61), linetype="dashed", size=0.8)+
  geom_segment(x=-45,y=2.747481,xend=61,yend=5.204455, color="black", size=1)+
  theme_bw()+ # now for some reason you need to put the theme_bw() ahead of the theme() or nothing in theme() will work
  theme(strip.text = element_text(size = 11))+
  annotate("text", x = -150, y = 4.9, label = "paste(italic(R) ^ 2, \" = 0.4874\")", parse = TRUE)+
  annotate("text", x = -150, y = 5.6, label = "paste(italic(p), \" = 5.14e-05\")", parse = TRUE)+
  annotate("rect", xmin = -225, xmax = -75, ymin = 4.5, ymax = 6, alpha = .2)

div.fig <- ggarrange(nssplot,ssplot, heights = c(1,1), ncol=1, nrow=2, labels = c("A", "B")) # Figure 5
div.fig <- saveRDS(div.fig, "figures/figure5") # export alpha diversity figure
```

## beta diversity

#### transforming ASV table
```{r}
# Hellinger transformation
otu.hel <- otu_table(decostand(otu_table(ps.frdp), method = "hellinger"), taxa_are_rows=FALSE)
ps.hel <- phyloseq(tax_table(ps.frdp),
                    sample_data(ps.frdp),
                    otu_table(otu.hel),
                    phy_tree(ps.frdp),
                    refseq(ps.frdp)) 
```

#### permanova tests
Testing the influence of stage (seep, non-steady-state, and steady-state) and geochemical zone (linear sulfate-reduction zone, non-steady-state reduction zone, and below-sulfate-methane-transition-zone) on community composition  
```{r}
sample_data(ps.hel)$smtpos <- ifelse(sample_data(ps.hel)$geochem_zone == "linear SR zone","above SMTZ",
                              ifelse(sample_data(ps.hel)$geochem_zone == "nonlinear SR zone","above SMTZ",
                              ifelse(sample_data(ps.hel)$geochem_zone == "below SMTZ", "below SMTZ", NA))) # writing above-below smtz as another categorical variable
metadata.ps.hel <- as(sample_data(ps.hel), "data.frame") #  create sample data dataframe

# binary jaccard
dm.jac <- phyloseq::distance(ps.frdp, method = "jaccard", binary = TRUE) # makes binary Jaccard distance matrix (no need to transform)
adonis(dm.jac ~ geochem_zone*stage, data=metadata.ps.hel)

# Bray-Curtis
dm.bc <- phyloseq::distance(ps.hel, method = "bray") # makes bray-curtis distance matrix
adonis(dm.bc ~ geochem_zone*stage, data=metadata.ps.hel)

# unweighted Unifrac
set.seed(1)
dm.hel.unifrac <- UniFrac(ps.hel, weighted = FALSE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # create weighted Unifrac distance matrix
adonis(dm.hel.unifrac ~ geochem_zone*stage, data=metadata.ps.hel) 

# weighted Unifrac 
set.seed(1)
dm.hel.wunifrac <- UniFrac(ps.hel, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # create weighted Unifrac distance matrix
adonis(dm.hel.wunifrac ~ geochem_zone*stage, data=metadata.ps.hel) # geochem zone is 9.6%

# other variables are all significant, but possibly redundant:
adonis(dm.hel.wunifrac ~ stage, data=metadata.ps.hel) #  stage alone is 19.9%
adonis(dm.hel.wunifrac ~ core_flowtype, data=metadata.ps.hel) # significant, but at 13% is less than stage (which it is redundant with)
adonis(dm.hel.wunifrac ~ SO4_mM, data=metadata.ps.hel) # SO4 more than geochem zone (maybe more helpful, too? but redundant)
adonis(dm.hel.wunifrac ~ HS_mM, data=metadata.ps.hel) # significant but redundant
adonis(dm.hel.wunifrac ~ Alk_mM, data=metadata.ps.hel) # significant but redundant
adonis(dm.hel.wunifrac ~ core, data=metadata.ps.hel) # core matters more than any other, but "core" itself is a meaningless variable
adonis(dm.hel.wunifrac ~ pingo, data=metadata.ps.hel) # gas hydrate mound the cores are sampled from matters more as well, but also a meaningless variable
adonis(dm.hel.wunifrac ~ smtzposition, data=metadata.ps.hel) # above/below SMT explains only 5%
adonis(dm.hel.wunifrac ~ geochem_zone*core, data=metadata.ps.hel) # no significant interaction between core and geochem zone
```
Using the weighted Unifrac distance, both geochem_zone & stage were significant (stage explaining 16-17% of variance, while zone 8-9%). Interactions were also significant at around 4.9%. This did not vary by the transformation type (I tried variance-stabilizing transformation as well and got nearly the same results). Bray-Curtis, unweighted Unifrac, and Jaccard distance metrics showed the same patterns, though they respectively explained decreasing amounts of variance overall. So it seems relative abundances are a bit more influential than taxonomy alone in explaining differences between these communities, though they both help.  

Pingo (which Gas Hydrate Mound a sample was collected from) explains a bit more variance than does stage, and Core does even more. But it is a confounding variable (each pingo is at a separate methane flux stage), and which particular pingo a sample comes from offers no real helpful information. Core also confounds with pingo, core_flowtype, stage, etc. Even though core was the single variable that explained more of the variance than any other, geochem_zone*stage explained slightly more. Furthermore, core and geochem_zone do not have a significant interaction.  

#### ordinations 
weighted Unifrac with Hellinger transformation  
```{r}
# make better labels for plotting
sample_data(ps.hel)$stage <- ifelse(sample_data(ps.hel)$stage == "seep", "active methane seepage",
                        ifelse(sample_data(ps.hel)$stage == "fluxincreasing", "increasing methane flux",
                        ifelse(sample_data(ps.hel)$stage == "steadystate", "steady-state", NA)))

sample_data(ps.hel)$geochem_zone <- ifelse(sample_data(ps.hel)$geochem_zone == "lin", "linear SR zone",
                        ifelse(sample_data(ps.hel)$geochem_zone == "nss", "nonlinear SR zone",
                        ifelse(sample_data(ps.hel)$geochem_zone == "below", "below SMTZ", NA)))

sample_data(ps.hel)$geochem_zone <- factor(sample_data(ps.hel)$geochem_zone, levels = c("linear SR zone", "nonlinear SR zone", "below SMTZ")) # reorder 

# trying NMDS and PCoAs
set.seed(1)
ord.ps.hel.wuni.nmds <- ordinate(ps.hel, "NMDS", "unifrac", weighted=TRUE) # Stress = 0.135
set.seed(1)
ord.ps.hel.wuni.pcoa <- ordinate(ps.hel, "PCoA", "unifrac", weighted=TRUE)

# plotting ordinations by stage 
lb1 <- paste("R^2 == 0.197")
ord1.p.stage <- plot_ordination(ps.hel, ord.ps.hel.wuni.pcoa, color = "stage")+
  stat_ellipse()+
  scale_color_manual("",values = c("#fc8d62", "#66c2a5", "#8da0cb"))+
  ggtitle("A", subtitle = "Cores classified by methane stage")+
  annotate(geom="text", x=-0.1, y=0.28, label="p < 0.001") +
  annotate(geom="text", x=-0.1, y=0.34, label=lb1, parse=TRUE)+
  theme_bw()+
  theme(legend.position = "bottom")+
  guides(col = guide_legend(ncol = 1))

# plotting ordinations by redox zone 
lb2 <- paste("R^2 == 0.096")
ord1.p.zone<- plot_ordination(ps.hel, ord.ps.hel.wuni.pcoa, color = "geochem_zone")+
  stat_ellipse()+
  scale_color_manual("",values = c("#e78ac3","#a6d854","#ffd92f"))+
  ggtitle("B", subtitle = "Samples classified by redox zone")+
  annotate(geom="text", x=-0.15, y=0.17, label="p < 0.001", parse=TRUE)+
  annotate(geom="text", x=-0.15, y=0.23, label=lb2, parse=TRUE)+
  theme_bw()+
  theme(legend.position = "bottom")+
  guides(col = guide_legend(ncol = 1))

# showing them both together
ord.fig <- ord1.p.stage + ord1.p.zone
ord.fig
ord.fig <- saveRDS(ord.fig, "figures/figure6") # export figure
```
The PCoA has fewer outliers, and explains 25% x 17.4% variance on the two major axes, so I'll stick with that over NMDS.  

Maybe don't even show the plot by geochem zone... it's just a mess that doesn't suggest any clear grouping pattern. Just showing it because it tested as significant doesn't seem a good enough reason...

Can also add the sections "more permanovas" from lines 844-920, and possibly beta-diversity from lines 938-974. Just to be thorough...   

## follow-up tests on beta-diversity
```{r eval=FALSE}
library(phyloseq)
# omitting below-SMT samples, is there still a difference by stage?
ps.hel.above <- subset_samples(ps.hel, geochem_zone!="below") # subset
set.seed(1)
dm.hel.above.wunifrac <- UniFrac(ps.hel.above, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.above <- as(sample_data(ps.hel.above), "data.frame") # write metadata data frame
adonis(dm.hel.above.wunifrac ~ stage, data=metadata.ps.hel.above) # yes there is a difference: R2, p are 0.21521  0.001 ***

# considering only below-SMT samples, is there still a difference by stage?
ps.hel.below <- subset_samples(ps.hel, geochem_zone=="below") # subset
set.seed(1)
dm.hel.below.wunifrac <- UniFrac(ps.hel.below, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.below <- as(sample_data(ps.hel.below), "data.frame") # write metadata data frame
adonis(dm.hel.below.wunifrac ~ stage, data=metadata.ps.hel.below) # yes there is a difference: R2, p are 0.18597  0.003 **

# differences between the two stages steady-state and fluxincreasing? 
ps.hel.gc <- subset_samples(ps.hel, stage!="seep") # subset
set.seed(1)
dm.hel.gc.wunifrac <- UniFrac(ps.hel.gc, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.gc <- as(sample_data(ps.hel.gc), "data.frame") # write metadata data frame
adonis(dm.hel.gc.wunifrac ~ stage, data=metadata.ps.hel.gc) # there is a difference: R2, p are 0.06055  0.001 ***

# differences between the two stages (steady-state and fluxincreasing) when only considering above-SMT samples? 
ps.hel.above.gc <- subset_samples(ps.hel.above, stage!="seep") # subset
set.seed(1)
dm.hel.above.gc.wunifrac <- UniFrac(ps.hel.above.gc, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.above.gc <- as(sample_data(ps.hel.above.gc), "data.frame") # write metadata data frame
adonis(dm.hel.above.gc.wunifrac ~ stage, data=metadata.ps.hel.above.gc) #  there is not really a difference: R2, p are 0.04448  0.056 .

# differences between linear SR and non-steady-state SR across both seep and fluxincreasing?
ps.hel.above.flux <- subset_samples(ps.hel, stage!="steadystate" & geochem_zone!="below") # subset
set.seed(1)
dm.hel.above.flux.wunifrac <- UniFrac(ps.hel.above.flux, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.above.flux <- as(sample_data(ps.hel.above.flux), "data.frame") # write metadata data frame
adonis(dm.hel.above.flux.wunifrac ~ geochem_zone, data=metadata.ps.hel.above.flux) #  there is a difference: R2, p are 0.16792  0.004 **

# differences between linear SR and non-steady-state SR across fluxincreasing only?
ps.hel.above.fluxinc <- subset_samples(ps.hel.above.flux, stage=="fluxincreasing") # subset
set.seed(1)
dm.hel.above.fluxinc.wunifrac <- UniFrac(ps.hel.above.fluxinc, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.above.fluxinc <- as(sample_data(ps.hel.above.fluxinc), "data.frame") # write metadata data frame
adonis(dm.hel.above.fluxinc.wunifrac ~ geochem_zone, data=metadata.ps.hel.above.fluxinc) #  there is barely a difference: R2, p are ~0.3, ~0.05 depending on set.seed

# differences between linear SR and non-steady-state SR across seep only?
ps.hel.above.fluxseep <- subset_samples(ps.hel.above.flux, stage=="seep") # subset
set.seed(1)
dm.hel.above.fluxseep.wunifrac <- UniFrac(ps.hel.above.fluxseep, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.above.fluxseep <- as(sample_data(ps.hel.above.fluxseep), "data.frame") # write metadata data frame
adonis(dm.hel.above.fluxseep.wunifrac ~ geochem_zone, data=metadata.ps.hel.above.fluxseep) #  there is a difference: R2, p are 0.34069  0.002 **

# differences between non-steady-state SR and below-SMT across fluxincreasing only?
ps.hel.fluxinc.lower <- subset_samples(ps.hel, stage=="fluxincreasing" & geochem_zone!="lin") # subset
set.seed(1)
dm.hel.fluxinc.lower.wunifrac <- UniFrac(ps.hel.fluxinc.lower, weighted = TRUE, normalized = TRUE,  parallel = FALSE, fast = TRUE) # run unifrac
metadata.ps.hel.fluxinc.lower <- as(sample_data(ps.hel.fluxinc.lower), "data.frame") # write metadata data frame
adonis(dm.hel.fluxinc.lower.wunifrac ~ geochem_zone, data=metadata.ps.hel.fluxinc.lower) #  there is no difference: R2, p are 0.12587  0.412 BUT ONLY 8 SAMPLES

# following this up with an ordination of fluxincreasing GC samples and color by geochem_zone
ps.hel.fluxinc <- subset_samples(ps.hel, stage=="fluxincreasing") # subset
ord.ps.helfluxinc.wuni.pcoa <- ordinate(ps.hel.fluxinc, "PCoA", "unifrac", weighted=TRUE) # ordinate
plot_ordination(ps.hel.fluxinc, ord.ps.helfluxinc.wuni.pcoa, color = "geochem_zone")+stat_ellipse() # from the very few samples we have here, the idea that the linear SR zone is the outlier is supported

# t-tests of beta-diversity between stages
ps.hel.seep <- subset_samples(ps.hel, stage=="seep") # subset
ps.hel.fluxinc
ps.hel.ss <- subset_samples(ps.hel, stage=="steadystate")

set.seed(1)
dm.seep <- UniFrac(ps.hel.seep, weighted=TRUE, normalized=TRUE, parallel=FALSE, fast=TRUE) # calculate distance matrices for beta-diversity calculations
set.seed(1)
dm.fluxinc <- UniFrac(ps.hel.fluxinc, weighted=TRUE, normalized=TRUE, parallel=FALSE, fast=TRUE)
set.seed(1)
dm.ss <- UniFrac(ps.hel.ss, weighted=TRUE, normalized=TRUE, parallel=FALSE, fast=TRUE)

bdiv.seep <- as.vector(dm.seep) # rewrite the triangle distance matrices as vectors
bdiv.fluxinc <- as.vector(dm.fluxinc) 
bdiv.ss <- as.vector(dm.ss) 

t.test(bdiv.seep, bdiv.ss) 
t.test(bdiv.fluxinc, bdiv.ss) 
t.test(bdiv.fluxinc, bdiv.seep) 
```

Conclusions: all stages are distinctly different from one another. This is true even when considering only above-SMTZ or below-SMTZ samples. There appear to be differences between steady-state and non-steady-state regions of the sulfate-reduction zone, though this difference was more evident in PC1029 samples than in the gravity cores. We saw no evidence of differences between non-steady-state SR zones and below-SMTZ zones in cores where flux was increasing, though this may be a result of inadequate sampling (only 8 samples tested). There is also evidence for lower beta-diversity among seep samples than among the other two stages, suggesting there is some aspect of community convergence (not that surprising).   

## differentially abundant ASVs across methane flux stages and above/below the SMTZ
We have 76 samples: 15 from seep, 49 from steady-state, and 12 from flux-increasing stages. There are no seep samples from below-SMTZ, 16 from steady-state, and 4 from flux-increasing. In an ideal world, we would have a more balanced sample design for assessing ASVs that are differentially abundant between these three stages (above the SMTZ) and between the latter two stages (below SMTZ). However, DESeq2 is able to handle imbalances in group numbers fairly robustly: https://support.bioconductor.org/p/115433/   

```{r fig.height=3.5}
# enter logical metadata categories corresponding to whether sample stage is seep or increasing flux or steady-state
sample_data(ps.frdp)$is.seep <- FALSE
sample_data(ps.frdp)[which(sample_data(ps.frdp)$stage=="seep"),22] <- TRUE
sample_data(ps.frdp)$is.flux <- FALSE
sample_data(ps.frdp)[which(sample_data(ps.frdp)$stage=="fluxincreasing"),23] <- TRUE
sample_data(ps.frdp)$is.ss <- FALSE
sample_data(ps.frdp)[which(sample_data(ps.frdp)$stage=="steadystate"),24] <- TRUE

# subset the phyloseq object by above/below SMTZ
ps.above <- subset_samples(ps.frdp, smtzposition=="above") 
ps.below <- subset_samples(ps.frdp, smtzposition=="below") 

# make the same phyloseq object, but otu table with relative abundance instead of raw counts
ps.above.ra <- transform_sample_counts(ps.above, function(OTU) OTU/sum(OTU))
ps.below.ra <- transform_sample_counts(ps.below, function(OTU) OTU/sum(OTU))

# define a function which will make a table out of deseq biomarker results, and return taxonomy and relative abundance info 
table.stage.from.deseq <- function(deseq, relabundps, group){
  de.results <- DESeq(deseq, test = "Wald", fitType = "parametric", sfType = "poscounts")
  de.results.table <- results(de.results, cooksCutoff = FALSE)
  de.results.table <- de.results.table[which(de.results.table$padj < 0.05), ]
  meanabund.vector <- vector("numeric", length(rownames(de.results.table)))
  for (i in rownames(de.results.table)) {meanabund.vector[[i]] <- mean(otu_table(relabundps)[which(sample_data(relabundps)$stage==group),i])}
  meanabund.vector <- meanabund.vector[(0.5*length(meanabund.vector)+1):length(meanabund.vector)] 
  de.results.table <- cbind(as(meanabund.vector, "matrix"), as(de.results.table, "data.frame"), as(tax_table(relabundps)[rownames(de.results.table), ], "matrix"))
  colnames(de.results.table)[1] <- "relabund"
  return(de.results.table)
}

# then convert phyloseq object to deseq, and run the function for each stage of methane dynamics
de.above.is.seep <- phyloseq_to_deseq2(ps.above, ~ is.seep)
de.results.table.above.seep <- table.stage.from.deseq(deseq = de.above.is.seep, relabundps = ps.above.ra, group = "seep") 
de.above.is.flux <- phyloseq_to_deseq2(ps.above, ~ is.flux)
de.results.table.above.flux <- table.stage.from.deseq(deseq = de.above.is.flux, relabundps = ps.above.ra, group = "fluxincreasing") 
de.above.is.ss <- phyloseq_to_deseq2(ps.above, ~ is.ss)
de.results.table.above.ss <- table.stage.from.deseq(deseq = de.above.is.ss, relabundps = ps.above.ra, group = "steadystate") 

de.below.is.flux <- phyloseq_to_deseq2(ps.below, ~ is.flux)
de.results.table.below.flux <- table.stage.from.deseq(deseq = de.below.is.flux, relabundps = ps.below.ra, group = "fluxincreasing") 
de.below.is.ss <- phyloseq_to_deseq2(ps.below, ~ is.ss)
de.results.table.below.ss <- table.stage.from.deseq(deseq = de.below.is.ss, relabundps = ps.below.ra, group = "steadystate") 

# assign category for each de.results.table
de.results.table.above.seep$de.group <- "above.seep"
de.results.table.above.flux$de.group <- "above.flux"
de.results.table.above.ss$de.group <- "above.ss"
de.results.table.below.flux$de.group <- "below.flux"
de.results.table.below.ss$de.group <- "below.ss"

# rewrite ASV numbers so rbind will not rename duplicates
de.results.table.above.seep$asv <- rownames(de.results.table.above.seep)
de.results.table.above.flux$asv <- rownames(de.results.table.above.flux)
de.results.table.above.ss$asv <- rownames(de.results.table.above.ss)
de.results.table.below.flux$asv <- rownames(de.results.table.below.flux)
de.results.table.below.ss$asv <- rownames(de.results.table.below.ss)

# merge the biomarkers from each block into a single data frame
all.biomarkers <- rbind(as(de.results.table.above.seep, "data.frame"), 
                        as(de.results.table.above.flux, "data.frame"), 
                        as(de.results.table.above.ss, "data.frame"),
                        as(de.results.table.below.flux, "data.frame"),
                        as(de.results.table.below.ss, "data.frame")) 

# select only the ones more abundant in the blocks (as opposed to less)
all.biomarkers <- subset(all.biomarkers, log2FoldChange>0) 

# add in metadata about SMT position only, and stage only
all.biomarkers$smtpos <- "above"
all.biomarkers[which(all.biomarkers$de.group=="below.flux"),16] <- "below"
all.biomarkers[which(all.biomarkers$de.group=="below.ss"),16] <- "below"
all.biomarkers$stage <- "increasing methane flux"
all.biomarkers[which(all.biomarkers$de.group=="above.seep"),17] <- "active methane seepage"
all.biomarkers[which(all.biomarkers$de.group=="above.ss"),17] <- "steady-state"
all.biomarkers[which(all.biomarkers$de.group=="below.ss"),17] <- "steady-state"

# subset all.biomarkers dataframe by categories we want to graph at
most.biomarkers <- all.biomarkers %>% filter(Class!="Deltaproteobacteria" & Class!="Methanomicrobia") # remove Deltaproteobacteria & ANMEs because we'll graph them at higher taxonomic resolution, also removes NAs
anme.biomarkers <- all.biomarkers %>% filter(Class == "Methanomicrobia") # subset ANMEs only (all Methanomicrobial biomarkers are ANME in this dataset)
delta.biomarkers <- all.biomarkers %>% filter(Class == "Deltaproteobacteria" & is.na(Genus)==FALSE) # subset Deltaproteobacteria only

# when plotting, facet by more than two stages: Plotting most biomarkers at the class level, but ANMEs at Family and Deltas at Genus
most.biomarkers$plotlevel <- "Class"
anme.biomarkers$plotlevel <- "Family"
delta.biomarkers$plotlevel <- "Genus"
most.biomarkers$taxlevel <- most.biomarkers$Class
anme.biomarkers$taxlevel <- anme.biomarkers$Family
delta.biomarkers$taxlevel <- delta.biomarkers$Genus
biomarkers.to.plot <- rbind(most.biomarkers, anme.biomarkers, delta.biomarkers) # combine all subsetted dataframes for plotting

gg.all.biom <- ggplot(biomarkers.to.plot, aes(x=factor(taxlevel, levels = rev(levels(factor(taxlevel)))), y=log2FoldChange, color=Kingdom)) + 
  geom_point(aes(size = relabund*100, fill=Kingdom), color="black", pch=21) +
  scale_size_area("Percent abundance", max_size = max(all.biomarkers$relabund)*500)+
  scale_fill_discrete("Domain")+
  scale_x_discrete("")+
  scale_y_continuous("log2 Differential Abundance Change", position = "right", limits = c(0,27)) +
  facet_grid(smtpos+plotlevel~stage, scales = "free", space = "free") +
  coord_flip() +
  theme_bw()+
  theme(strip.text.x = element_text(size=11), strip.text.y = element_text(size=8),
        axis.text.x = element_text(size=10))
gg.all.biom
gg.all.biom <- saveRDS(gg.all.biom, "figures/figure7") # export figure
```
