---
title: "Supplemental figures and tables"
author: "Scott Klasek"
date: "6/10/2021"
output: github_document
---
Supplemental figures S5-S8 and tables S1 & S2 are presented here. Other figures/tables include:  

Figure S1- modeling setup  
Figure S2- porewater ammonium concentrations and expected POCSR from Storfjordrenna cores described in this study and elsewhere  
Figure S3- microbial community and porewater data from GC1048, which is shown [here](https://github.com/sklasek/svalflux/blob/master/markdowns/3_core_info_and_panel_figures.md)   
Figure S4- siboglinid tube pictures from cores at GHM3 seep site  
Table S3- compilation of porewater ammonium concentration data from Storfjordrenna GHMs  

## load necessary libraries and data
```{r}
library(phyloseq)
library(vegan)
library(DESeq2)
library(tidyverse)
library(patchwork)
library(ggpmisc)
library(tables)
library(here)
sessioninfo <- sessionInfo()
ps.frdp <- readRDS(file="/Users/scottklasek/Desktop/svalflux/data/ps.frdp") # imports the final phyloseq object
```

## Fig. S5, regression of mcrA counts vs modeled AOM rates
I have omitted GC1048 samples from here.  
```{r}
mcra.aom <- read.csv(file="/Users/scottklasek/Desktop/svalflux/data/mcra.aomrate.2021.csv") # import csv
mcra.aom$logaom <- log10(mcra.aom$aom) # calculate log of AOM rates
mcra.aom$logmcra <- log10(mcra.aom$mcra) # calculate log of mcrA gene abundances

# assign stages or whatever we end up calling them
mcra.aom$stage <- "Steady-state"
mcra.aom[which(mcra.aom$core=="GC1045"),"stage"] <- "Non-steady-state"
mcra.aom[which(mcra.aom$core=="GC1081"),"stage"] <- "Non-steady-state"

mcra.aom.no48 <- mcra.aom %>% filter(core !="GC1048") # remove GC1048

# linear model steady-state
ss.lm <- lm(mcra.aom.no48 %>% filter(core_flowtype=="ss") %>% pull(logmcra) ~ 
           mcra.aom.no48 %>% filter(core_flowtype=="ss") %>% pull(logaom))  # steady-state regression
summary(ss.lm)

# plot steady-state
ss.df <- mcra.aom.no48 %>% filter(core_flowtype=="ss")
ss.mcra <- ggplot(ss.df, aes(logaom,logmcra,color=stage))+
  geom_point(size=2)+
  geom_smooth(method = "lm")+
  scale_color_manual("", values = "#8da0cb")+
  scale_x_continuous(expression('log'[10]*" AOM rate (µmols L"^{-1}*" day"^{-1}*")"))+
  scale_y_continuous(expression('log'[10]*' mcrA gene copies g'^{"-1"}), limits = c(3.5, 7.5))+
  theme(axis.title=element_text(size = 12), axis.text = element_text(size = 11),
        legend.title = element_text(size = 12), legend.text = element_text(size = 10))+
  stat_poly_eq(formula = ss.df$logmcra ~ ss.df$logaom, 
                aes(label = paste(..eq.label.., ..rr.label.., stat(p.value.label), sep = "~~~")), 
                parse = TRUE) +  
  theme_bw()+theme(legend.position = "bottom")+ggtitle("A")

# linear model non-steady-state
nss.lm <- lm(mcra.aom.no48 %>% filter(core_flowtype=="inc") %>% pull(logmcra) ~ 
           mcra.aom.no48 %>% filter(core_flowtype=="inc") %>% pull(logaom))  # non-steady-state regression
summary(nss.lm)

# plot non-steady-state
nss.df <- mcra.aom.no48 %>% filter(core_flowtype=="inc")
nss.mcra <- ggplot(nss.df, aes(logaom,logmcra,color=stage))+
  geom_point(size=2)+
  geom_smooth(method = "lm")+
  scale_color_manual("", values = "#66c2a5")+
  scale_x_continuous(expression('log'[10]*" AOM rate (µmols L"^{-1}*" day"^{-1}*")"))+
  scale_y_continuous(expression('log'[10]*' mcrA gene copies g'^{"-1"}), limits = c(3.5, 7.5))+
  theme(axis.title=element_text(size = 12), axis.text = element_text(size = 11),
        legend.title = element_text(size = 12), legend.text = element_text(size = 10))+
  stat_poly_eq(formula = nss.df$logmcra ~ nss.df$logaom, 
                aes(label = paste(..eq.label.., ..rr.label.., stat(p.value.label), sep = "~~~")), 
                parse = TRUE) +  
  theme_bw()+theme(legend.position = "bottom")+ggtitle("B")

ss.mcra + nss.mcra
```
Figure S3. Regression of mcrA gene copy numbers (per gram bulk sediment) to modeled AOM rates shows a linear relationship across samples from all cores (log-log transformation). Samples are colored according to stages of methane dynamics, and those that did not contain detectable mcrA were omitted.   

## Fig. S6 shows differentially abundant ASVs across methane flux stages and above/below the SMTZ
We have 63 samples: 15 from seep, 36 from steady-state, and 12 from flux-increasing stages. There are no seep samples from below-SMTZ, 15 from steady-state, and 4 from flux-increasing. In an ideal world, we would have a more balanced sample design for assessing ASVs that are differentially abundant between these three stages (above the SMTZ) and between the latter two stages (below SMTZ). However, DESeq2 is able to handle imbalances in group numbers fairly robustly: https://support.bioconductor.org/p/115433/   

```{r, fig.height=4}
# enter logical metadata categories corresponding to whether sample stage is seep or increasing flux or steady-state

sample_data(ps.frdp)$is.seep <- FALSE
sample_data(ps.frdp)[which(sample_data(ps.frdp)$stage=="seep"),22] <- TRUE
sample_data(ps.frdp)$is.flux <- FALSE
sample_data(ps.frdp)[which(sample_data(ps.frdp)$stage=="fluxincreasing"),23] <- TRUE
sample_data(ps.frdp)$is.ss <- FALSE
sample_data(ps.frdp)[which(sample_data(ps.frdp)$stage=="steadystate"),24] <- TRUE

# here is where I omit GC1048
ps.frdp.no48 <- subset_samples(ps.frdp, core!="GC1048")
ps.frdp.no48
# subset the phyloseq object by above/below SMTZ
ps.above <- subset_samples(ps.frdp.no48, smtzposition=="above") 
ps.below <- subset_samples(ps.frdp.no48, smtzposition=="below") 

# make the same phyloseq object, but otu table with relative abundance instead of raw counts
ps.above.ra <- transform_sample_counts(ps.above, function(OTU) OTU/sum(OTU))
ps.below.ra <- transform_sample_counts(ps.below, function(OTU) OTU/sum(OTU))

# define a function which will make a table out of deseq biomarker results, and return taxonomy and relative abundance info 
table.stage.from.deseq <- function(deseq, relabundps, group){
  de.results <- DESeq(deseq, test = "Wald", fitType = "parametric", sfType = "poscounts")
  de.results.table <- results(de.results, cooksCutoff = FALSE)
  de.results.table <- de.results.table[which(de.results.table$padj < 0.05), ]
  meanabund.vector <- vector("numeric", length(rownames(de.results.table)))
  for (i in rownames(de.results.table)) {meanabund.vector[[i]] <- mean(otu_table(relabundps)[which(sample_data(relabundps)$stage==group),i])}
  meanabund.vector <- meanabund.vector[(0.5*length(meanabund.vector)+1):length(meanabund.vector)] 
  de.results.table <- cbind(as(meanabund.vector, "matrix"), as(de.results.table, "data.frame"), as(tax_table(relabundps)[rownames(de.results.table), ], "matrix"))
  colnames(de.results.table)[1] <- "relabund"
  return(de.results.table)
}

# then convert phyloseq object to deseq, and run the function for each stage of methane dynamics
de.above.is.seep <- phyloseq_to_deseq2(ps.above, ~ is.seep)
de.results.table.above.seep <- table.stage.from.deseq(deseq = de.above.is.seep, relabundps = ps.above.ra, group = "seep") 
de.above.is.flux <- phyloseq_to_deseq2(ps.above, ~ is.flux)
de.results.table.above.flux <- table.stage.from.deseq(deseq = de.above.is.flux, relabundps = ps.above.ra, group = "fluxincreasing") 
de.above.is.ss <- phyloseq_to_deseq2(ps.above, ~ is.ss)
de.results.table.above.ss <- table.stage.from.deseq(deseq = de.above.is.ss, relabundps = ps.above.ra, group = "steadystate") 

de.below.is.flux <- phyloseq_to_deseq2(ps.below, ~ is.flux)
de.results.table.below.flux <- table.stage.from.deseq(deseq = de.below.is.flux, relabundps = ps.below.ra, group = "fluxincreasing") 
de.below.is.ss <- phyloseq_to_deseq2(ps.below, ~ is.ss)
de.results.table.below.ss <- table.stage.from.deseq(deseq = de.below.is.ss, relabundps = ps.below.ra, group = "steadystate") 

# assign category for each de.results.table
de.results.table.above.seep$de.group <- "above.seep"
de.results.table.above.flux$de.group <- "above.flux"
de.results.table.above.ss$de.group <- "above.ss"
de.results.table.below.flux$de.group <- "below.flux"
de.results.table.below.ss$de.group <- "below.ss"

# rewrite ASV numbers so rbind will not rename duplicates
de.results.table.above.seep$asv <- rownames(de.results.table.above.seep)
de.results.table.above.flux$asv <- rownames(de.results.table.above.flux)
de.results.table.above.ss$asv <- rownames(de.results.table.above.ss)
de.results.table.below.flux$asv <- rownames(de.results.table.below.flux)
de.results.table.below.ss$asv <- rownames(de.results.table.below.ss)

# merge the biomarkers from each block into a single data frame
all.biomarkers <- rbind(as(de.results.table.above.seep, "data.frame"), 
                        as(de.results.table.above.flux, "data.frame"), 
                        as(de.results.table.above.ss, "data.frame"),
                        as(de.results.table.below.flux, "data.frame"),
                        as(de.results.table.below.ss, "data.frame")) 

# select only the ones more abundant in the blocks (as opposed to less)
all.biomarkers <- subset(all.biomarkers, log2FoldChange>0) 

# add in metadata about SMT position only, and stage only
all.biomarkers$smtpos <- "above"
all.biomarkers[which(all.biomarkers$de.group=="below.flux"),16] <- "below"
all.biomarkers[which(all.biomarkers$de.group=="below.ss"),16] <- "below"
all.biomarkers$stage <- "increasing methane flux"
all.biomarkers[which(all.biomarkers$de.group=="above.seep"),17] <- "active methane seepage"
all.biomarkers[which(all.biomarkers$de.group=="above.ss"),17] <- "steady-state"
all.biomarkers[which(all.biomarkers$de.group=="below.ss"),17] <- "steady-state"

# subset all.biomarkers dataframe by categories we want to graph at
most.biomarkers <- all.biomarkers %>% filter(Class!="Deltaproteobacteria" & Class!="Methanomicrobia") # remove Deltaproteobacteria & ANMEs because we'll graph them at higher taxonomic resolution, also removes NAs
anme.biomarkers <- all.biomarkers %>% filter(Class == "Methanomicrobia") # subset ANMEs only (all Methanomicrobial biomarkers are ANME in this dataset)
delta.biomarkers <- all.biomarkers %>% filter(Class == "Deltaproteobacteria" & is.na(Genus)==FALSE) # subset Deltaproteobacteria only

# when plotting, facet by more than two stages: Plotting most biomarkers at the class level, but ANMEs at Family and Deltas at Genus
most.biomarkers$plotlevel <- "Community"
anme.biomarkers$plotlevel <- "ANME"
delta.biomarkers$plotlevel <- "SRB"
most.biomarkers$taxlevel <- most.biomarkers$Class
anme.biomarkers$taxlevel <- anme.biomarkers$Family
delta.biomarkers$taxlevel <- delta.biomarkers$Genus
biomarkers.to.plot <- rbind(most.biomarkers, anme.biomarkers, delta.biomarkers) # combine all subsetted dataframes for plotting

gg.all.biom <- ggplot(biomarkers.to.plot, aes(x=factor(taxlevel, levels = rev(levels(factor(taxlevel)))), y=log2FoldChange, color=Kingdom)) + 
  geom_point(aes(size = relabund*100, fill=Kingdom), color="black", pch=21) +
  scale_size_area("Percent abundance", max_size = max(all.biomarkers$relabund)*700, breaks=c(0.1, 0.3, 1, 3))+
  scale_fill_discrete("Domain")+
  scale_x_discrete("")+
  scale_y_continuous("log2 Differential Abundance Change", position = "right", limits = c(0,27)) +
  facet_grid(smtpos+plotlevel~stage, scales = "free", space = "free") +
  coord_flip() +
  theme_bw()+
  theme(strip.text.x = element_text(size=11), strip.text.y = element_text(size=8),
        axis.text.x = element_text(size=10), legend.position = "bottom", legend.box = "vertical")
gg.all.biom # export dimensions 6.3 x 6.9

biomarkers.to.plot %>% filter(Phylum=="Atribacteria") %>% select(Class, stage) # shows biomarker ASVs for different stages and positions relative to SMT
length(which(tax_table(ps.frdp)[,3]=="Calditrichia")) # shows numbers ASVs from a certain taxa

# how many ASVs are differentially abundant? what average %s are they of communities across stages?
length(unique(biomarkers.to.plot$asv))
sum(biomarkers.to.plot %>% filter(stage=="active methane seepage") %>% select(relabund))*100 # 17.4%
sum(biomarkers.to.plot %>% filter(stage=="steady-state") %>% select(relabund))*100 # 6.9%
sum(biomarkers.to.plot %>% filter(stage=="increasing methane flux") %>% select(relabund))*100 # 1.6%

# how many total ANME ASVs in the dataset?
nrow(data.frame(tax_table(ps.frdp.no48)) %>% filter(Order=="ANME-1" | Family=="ANME-2a-2b"| Family=="ANME-2c" | Genus=="ANME-3"))
```
(Made this figure a little easier to read in Affinity Designer).  

## Fig. S7, ordination of just increasing-flux cores by geochem_zone
```{r}
# transform the phyloseq object
otu.hel <- otu_table(decostand(otu_table(ps.frdp), method = "hellinger"), taxa_are_rows=FALSE)
ps.hel <- phyloseq(tax_table(ps.frdp),
                    sample_data(ps.frdp),
                    otu_table(otu.hel),
                    phy_tree(ps.frdp),
                    refseq(ps.frdp)) 

# make better labels for plotting
sample_data(ps.hel)$stage <- ifelse(sample_data(ps.hel)$stage == "seep", "active methane seepage",
                        ifelse(sample_data(ps.hel)$stage == "fluxincreasing", "increasing methane flux",
                        ifelse(sample_data(ps.hel)$stage == "steadystate", "steady-state", NA)))
sample_data(ps.hel)$geochem_zone <- ifelse(sample_data(ps.hel)$geochem_zone == "lin", "linear SR zone",
                        ifelse(sample_data(ps.hel)$geochem_zone == "nss", "nonlinear SR zone",
                        ifelse(sample_data(ps.hel)$geochem_zone == "below", "below SMT", NA)))
sample_data(ps.hel)$geochem_zone <- factor(sample_data(ps.hel)$geochem_zone, levels = c("linear SR zone", "nonlinear SR zone", "below SMT")) 

ps.hel.fluxinc <- subset_samples(ps.hel, stage=="increasing methane flux") # subset for samples experiencing methane flux increase
ord.ps.helfluxinc.wuni.pcoa <- ordinate(ps.hel.fluxinc, "PCoA", "unifrac", weighted=TRUE) # ordinate

fluxord <- plot_ordination(ps.hel.fluxinc, ord.ps.helfluxinc.wuni.pcoa, color = "geochem_zone")+
  scale_color_manual("Redox zones",values = c("#e78ac3","#a6d854","#ffd92f"))+
  stat_ellipse()+
  theme_bw()
fluxord 
```
From the very few samples we have here, the idea that the linear SR zone is the outlier is supported.   

## Figure S8 [network here](https://github.com/sklasek/svalflux/blob/master/markdowns/7_network.md).  

## Table S1: table of core info
```{r}
coreinfo <- read.csv(file="/Users/scottklasek/Desktop/svalflux/data/coreinfo.csv") # import csv
names(coreinfo)[1:6] <- c("Core", "Latitude", "Longitude", "Water depth (m)", "Core recovery (cm)", "SMT depth (cm)")
flux <- c("~100 (estimated, see text)", 5.28, 4.22, 0.41, 1.30, 1.30, 1.30)
coreinfo <- cbind(coreinfo, flux)
colnames(coreinfo)[7] <- "Methane flux (mol m-2 yr-1) at time of sampling"
coreinfo
```
Table S1. Latitude, longitude, water depth, core recovery, and sulfate-methane transition depth of all cores analyzed in this study.   

## Table S2: methane flux histories 
```{r}
inc.core <- read.csv(file="/Users/scottklasek/Desktop/svalflux/data/increasing.core.flux.table.csv") # import csv
names(inc.core)[2:4] <- c("Year before sampling", "CH4 flux mols m^-2 yr ^-1", "Peak AOM depth (cm)") # rename columns to be more legible
inc.core
```
Table S2. Increases in methane flux over the past two decades for cores GC1045 and GC1081 and corresponding depths of modeled peak AOM rates. Fluxes are integrated from AOM rate data, using cell widths of 2.5 cm. (Peak AOM depths also at 2.5 cm resolution).   

## session info
```{r}
sessioninfo
```

