---
title: "bubble_plots_8.20"
author: "Scott Klasek"
date: "8/26/2020"
output: github_document
---

## recreate bubble plots from the phyloseq object

## load necessary libraries
```{r}
library(tidyverse)
library(phyloseq)
library(here)
sessioninfo <- sessionInfo()
```

## import phyloseq object
ps.frdp is from sequence_processing Rmd script  
also calculate a few stats of abundant ASVs
```{r}
ps.frdp <- readRDS(file="ps.frdp") # imports the final phyloseq object

# what are the most abundant ASVs?
ps.ra <- transform_sample_counts(ps.frdp, function(OTU) OTU/sum(OTU))
asv.ras <- colSums(otu_table(ps.ra)*100/(nsamples(ps.ra))) # calculates the average percent abundance of each ASV in dataset
asvpcts <- data.frame(cbind(names(asv.ras), asv.ras, tax_table(ps.ra))) # combine taxonomy and percent abundances of most abundant ASVs
colnames(asvpcts)[1] <- "ASV" # rename column
colnames(asvpcts)[2] <- "percent_abundance" # rename column
asvpcts$percent_abundance <- as.numeric(as.character(asvpcts$percent_abundance)) # convert to numeric
asvpcts.abund <- asvpcts %>% filter(percent_abundance > 1) # select only the ones above 1% 
asvpcts.abund

# how abundant are ANME and SRB?
asvANMEs <- asvpcts %>% filter(Family=="ANME-1a" | Family=="ANME-1b" | Family=="ANME-2a-2b" | Family=="ANME-2c")
sum(asvANMEs$percent_abundance) # ANME at 10.0%
asvSRBs <- asvpcts %>% filter(Genus=="SEEP-SRB1" | Genus=="SEEP-SRB2" | Genus=="Desulfatiglans" | Genus=="Desulfobulbus" | Genus=="Desulfococcus")
sum(asvSRBs$percent_abundance) # SRB at 12.0%
```

## Pull out and subset taxonomic and relative abundance data
Find the most abundant taxonomic classes, pull out their relative abundances, create a dataframe of them. Then identify the most abundant ANME and SRB and do the same thing. Add metadata and finally subset by stage.   
```{r}
classes.ps.frdp <- unique(tax_table(ps.frdp)[,3]) # obtains the unique class names
classes.ps.frdp <- classes.ps.frdp[!is.na(classes.ps.frdp)] # removes the NA

mx.f.asv <- as.matrix(otu_table(ps.frdp)) # creates a matrix  from the phyloseq object
class.ra <- vector("list",0)
for (i in classes.ps.frdp) {class.ra[[i]] <- sum(mx.f.asv[,which(tax_table(ps.frdp)[,3]==i)])/(sum(otu_table(ps.frdp))/100)}
class.output <- cbind(class.ra) 
# cab print list of classes by their relative abundance: print(class.output) 
top.c <- c("JS1","Deltaproteobacteria","Methanomicrobia","Dehalococcoidia","Anaerolineae","Campylobacteria","Aminicenantia","Bacteroidia","Phycisphaerae","Lokiarchaeia","Woesearchaeia","Gammaproteobacteria","Thermoplasmata","Alphaproteobacteria", "Nitrososphaeria") # all classes > 1% relabund
sum(25.57895,14.68791,10.22672,7.295351,4.331321,3.897544,3.558376,2.645234,2.289897,2.236513,1.741267,1.636488,1.2688,1.181987,1.01452) # 83.6% of sequences belong to these 15 classes. Sum of all 125 is probably 90.% (prior run flux_dada2)

f.ra <- transform_sample_counts(ps.frdp, function(OTU) OTU/sum(OTU)) # creates relative abundance phyloseq object
f.ra.top.c <- subset_taxa(f.ra, Class=="JS1" | Class=="Deltaproteobacteria" | Class=="Methanomicrobia" | Class=="Dehalococcoidia" | Class=="Anaerolineae" | Class=="Campylobacteria" | Class=="Aminicenantia" | Class=="Bacteroidia" | Class=="Phycisphaerae" | Class=="Lokiarchaeia" | Class=="Woesearchaeia" | Class=="Gammaproteobacteria" | Class=="Thermoplasmata" | Class=="Alphaproteobacteria" | Class=="Nitrososphaeria") # top 15 classes

# use a nested for-loop to pull relative abundance data out of phyloseq objects (by taxonomic division):
graphclass <- unique(tax_table(f.ra.top.c)[,3])
graphclass <- as.vector(graphclass)
class.sums.m <- matrix(nrow=76, ncol=15) # define matrix dimensions like this
rownames(class.sums.m) <- sample_names(f.ra.top.c) # specify row names
colnames(class.sums.m) <- graphclass # specify column names

for (i in sample_names(f.ra.top.c)) { 
  for (j in graphclass) {
        class.sums.m[[i,j]] <- sum(otu_table(f.ra.top.c)[sample_names(f.ra.top.c)==i,which(tax_table(f.ra.top.c)[,3]==j)]) # this is calculating the average percent abundance of each class
    }
}

# make data frames (in long format) from the matrix of abundances
flux.class.wide <- t(class.sums.m)
taxonrep <- rep(rownames(flux.class.wide), each=76)
tax <- as.character(taxonrep)
grouprep <- colnames(flux.class.wide)
groups <- rep(grouprep,times=15)
abund <- as.numeric(t(flux.class.wide))
longdf <- cbind.data.frame(tax,abund,groups)

# ANME and SRB info
# manually inspected taxonomy at different levels to decide which ANME families and SRB genera to include on the graph using variations of the commands directly below:
methanoASVs <- which(tax_table(f.ra)[,3]=="Methanomicrobia") # lists the ASVs that belong to Class Methanomicrobia
unique(tax_table(f.ra)[methanoASVs,5]) # shows family-level taxonomy of methano ASVs, see the top one is an ANME-1a
sum(otu_table(f.ra)[,which(tax_table(f.ra)[,5]=="ANME-1a")])/.76 # ANME-1a are 4.55% of sequences

# here's who I chose
ANMEs <- c("ANME-1a","ANME-1b","ANME-2c","ANME-2a-2b") # selected ANMEs at the Family level (Genus level contains many NAs)
SRBs <- c("SEEP-SRB1","SEEP-SRB2","Desulfatiglans","Desulfococcus","Desulfobulbus") # selected SRBs at the Genus level

flux.srbs <- subset_taxa(f.ra, Genus=="SEEP-SRB1" | Genus=="SEEP-SRB2" | Genus=="Desulfatiglans"| Genus=="Desulfococcus"| Genus=="Desulfobulbus" ) # contains 271 ASVs
flux.anme <- subset_taxa(f.ra, Family=="ANME-1a" | Family=="ANME-1b" | Family=="ANME-2a-2b" | Family=="ANME-2c") # contains 36 ASVs

# Pull out abundance data for ANMEs 
graphanme <- unique(tax_table(flux.anme)[,5])
graphanme <- as.vector(graphanme)
family.sums.m <- matrix(nrow=76, ncol=4) # define matrix dimensions 
rownames(family.sums.m) <- sample_names(flux.anme) # specify row names
colnames(family.sums.m) <- graphanme # specify column names. matrix is 76 x 4

for (i in sample_names(flux.anme)) { 
  for (j in graphanme) {
        family.sums.m[[i,j]] <- sum(otu_table(flux.anme)[sample_names(flux.anme)==i,which(tax_table(flux.anme)[,5]==j)]) # this is calculating the average percent abundance of each ANME family
    }
}

# Pull out abundance data for SRBs
graphsrbs <- unique(tax_table(flux.srbs)[,6])
graphsrbs <- as.vector(graphsrbs)
gen.sums.m <- matrix(nrow=76, ncol=5) # define matrix dimensions 
rownames(gen.sums.m) <- sample_names(flux.srbs) # specify row names
colnames(gen.sums.m) <- graphsrbs # specify column names

for (i in sample_names(flux.srbs)) { 
  for (j in graphsrbs) {
        gen.sums.m[[i,j]] <- sum(otu_table(flux.srbs)[sample_names(flux.srbs)==i,which(tax_table(flux.srbs)[,6]==j)]) # this is calculating the average percent abundance of each SRB genus
    }
}

# now turn these matrices into long dataframe form, combine them, and add metadata
# first for ANME
anme.taxonrep <- rep(colnames(family.sums.m), each=76)
anme.tax <- as.character(anme.taxonrep)
anme.grouprep <- rownames(family.sums.m)
anme.groups <- rep(anme.grouprep,times=4)
anme.abund <- as.numeric(family.sums.m)
longdf.family <- cbind.data.frame(anme.tax,anme.abund,anme.groups)
longdf.family$level <- "ANME"
longdf.family$Domain <- "Archaea"
longdf.family$tax_reorder <- factor(longdf.family$anme.tax, levels=c("ANME-1a","ANME-1b","ANME-2a-2b","ANME-2c"))

# now SRB
srb.taxonrep <- rep(colnames(gen.sums.m), each=76)
srb.tax <- as.character(srb.taxonrep)
srb.grouprep <- rownames(gen.sums.m)
srb.groups <- rep(srb.grouprep,times=5)
srb.abund <- as.numeric(gen.sums.m)
longdf.genus <- cbind.data.frame(srb.tax,srb.abund,srb.groups)
longdf.genus$level <- "SRB"
longdf.genus$Domain <- "Bacteria"
longdf.genus$tax_reorder <- factor(longdf.genus$srb.tax, levels = c("SEEP-SRB1","SEEP-SRB2","Desulfatiglans","Desulfococcus","Desulfobulbus"))

# combine all the dataframes
# first longdf, 1140 rows
longdf$level <- "Community"
longdf$Domain <- "Bacteria" # writes default domain as Bacteria
archaearows <- which(longdf$tax=="Methanomicrobia"|longdf$tax=="Woesearchaeia"|longdf$tax=="Lokiarchaeia"|longdf$tax=="Nitrososphaeria") # specifies the rows where classes belong to Archaea.
longdf[archaearows,5] <- "Archaea" # overwrites domain from Bacteria to Archaea
longdf$tax_reorder <- factor(longdf$tax, levels = c("JS1","Dehalococcoidia","Anaerolineae","Aminicenantia","Campylobacteria","Bacteroidia","Phycisphaerae","Lokiarchaeia","Woesearchaeia","Gammaproteobacteria","Thermoplasmata","Alphaproteobacteria","Nitrososphaeria","Methanomicrobia","Deltaproteobacteria"))

colnames(longdf.genus)[1:3] <- c("tax", "abund", "groups") # rewrites column names so that they are not specific to SRB
colnames(longdf.family)[1:3] <- c("tax", "abund", "groups") # rewrites column names so that they are not specific to ANME
superdf2 <- rbind(longdf,longdf.family,longdf.genus)
superdf2$pctabund <- 100*superdf2$abund

# add in metadata
repmeta <- do.call("rbind", replicate(24, sample_data(f.ra.top.c), simplify = FALSE)) # repeats metadata for each sample to match the rows in the long dataframe. Use the number 24 because there are 24 total taxonomies to plot amongst three taxonomic groups (class, family, genus)
repmeta$groups <- row.names(repmeta)
metasuperdf <- merge.data.frame(repmeta,superdf2) # merges long data frame with abundances with the metadata
metasuperdf$level_reorder <- factor(metasuperdf$level, levels=c("Community","ANME","SRB"))

# subset by core flowtype 
sdf1029 <- subset(metasuperdf, metasuperdf$stage=="seep") 
sdf.fi <- subset(metasuperdf, metasuperdf$stage=="fluxincreasing") 
sdf.ss <- subset(metasuperdf, metasuperdf$stage=="steadystate")


# stats calculating ANME abundances in total communities
anme76 <- sdf.fi %>% filter(level=="ANME" & depth==76)
anme110 <- sdf.fi %>% filter(level=="ANME" & depth==110)
srb76 <- sdf.fi %>% filter(level=="SRB" & depth==76)
srb110 <- sdf.fi %>% filter(level=="SRB" & depth==110)

sample_data(ps.frdp)

sdf.fi %>% filter(level=="ANME" & tax_reorder=="ANME-1b" & core=="GC1081" & depth==69)
```

## Graph community composition bubble plots
Note that the Classes were all mixed up when I was plotting tax_reorder under aes(). Replotting by tax solves this, but I wanted to order Deltaproteobacteria and Methanomicrobia next to ANME and SRB. It's a small detail though.  
```{r}
# first the seep site
bp1029 <- ggplot(sdf1029,aes(tax,depth,size=pctabund))
bp.seep <- bp1029+
  geom_point(aes(fill=Domain),colour="black",pch=21)+
  xlab("")+
  scale_y_reverse("", limits=c(35,0), breaks=c(0,5,10,15,20,25,30,35))+
  scale_fill_discrete("")+
  facet_grid(~level_reorder, scales = "free", space = "free")+
  scale_size_area("% Abundance",breaks=c(1,5,10,20,40,60))+
  theme_bw()+
  theme(legend.position = "top", legend.box = "vertical", axis.text.x=element_text(angle = 55, hjust = 1,      
    size=11), axis.text.y=element_text(size = 12), legend.text=element_text(size=11),
    axis.title=element_text(size=14), strip.text = element_text(size=11), plot.margin = margin(0,0,0,-0.5,"cm"))
bp.seep
bp.seep <- saveRDS(bp.seep, "figures/bp.seep") # export plot

# non-steady-state
bp.fi <- ggplot(sdf.fi,aes(tax,depth,size=pctabund))
bp.fluxincreasing <- bp.fi+
  geom_point(aes(fill=Domain),colour="black",pch=21)+
  xlab("")+
  facet_grid(core~level_reorder,scales="free", space = "free")+
  scale_y_reverse("",breaks=c(0,25,50,75,100,125,150),limits=c(150,0))+
  scale_fill_discrete("")+
  scale_size_area("% Abundance",breaks=c(1,5,10,20,40,60))+
  geom_hline(aes(yintercept = smt),linetype="dashed",size=0.8)+ # this line returns Error in -x : invalid argument to unary operator
  theme_bw()+
  theme(legend.position = "top", 
        axis.text.x=element_text(angle = 55, hjust = 1, size=12),
        axis.text = element_text(size=13),
        legend.text=element_text(size=13),
        strip.text.x = element_text(size=13), 
        strip.text.y = element_blank(), 
        plot.margin = margin(0,0,0,-0.25,"cm"), legend.box = "vertical")
bp.fluxincreasing 
bp.fluxincreasing <- saveRDS(bp.fluxincreasing, "figures/bp.fluxincreasing") # export plot

# steady-state
bp.ss <- ggplot(sdf.ss,aes(tax,depth,size=pctabund))
bp.steadystate <- bp.ss+
  geom_point(aes(fill=Domain),colour="black",pch=21)+
  xlab("")+
  facet_grid(core~level_reorder,scales="free", space = "free")+
  scale_y_reverse("",breaks=c(0,50,100,150,200,250,300,350),limits=c(350,0))+
  scale_fill_discrete("")+
  scale_size_area("% Abundance",breaks=c(1,5,10,20,40,60))+
  geom_hline(aes(yintercept = smt),linetype="dashed",size=0.8)+
  theme_bw()+
  theme(legend.position = "top",
        axis.text.x=element_text(angle = 55, hjust = 1, size=12),
        axis.text.y = element_text(size=13),
        legend.text=element_text(size=13),
        strip.text.x = element_text(size=13), 
        strip.text.y = element_blank(), 
        legend.box = "vertical", plot.margin = margin(0,0,0,-0.25,"cm"))
bp.steadystate
bp.steadystate <- saveRDS(bp.steadystate, "figures/bp.steadystate") # export plot
```


